Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-08 paper/main.tex
some edits to Section 3

For each cosmology, Mira-Titan contains a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$). 
These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), we see each low-resolution and the high-resolution run oscillate around 
a mean term, which is relatively smooth, compared to these realizations. When this
mean term is subtracted, we will obtain a plot like the right panel in 
Figure \ref{fig:plot_data}.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  $\bar{Y}$ is typically pre-processed
in a domain-specific manner, such as that defined in Section \ref{sec:data}.
Together, $X$ and $\bar{Y}$ comprise the training data for a particular cosmology.
We will reserve specific integration 
of expert knowledge for the CAMB and Mira-Titan suites, such as this pre-processing,
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

@@

Let $S$ represent the true (but unknown) underlying power matter spectra for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e. 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).

We next place a Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\boldsymbol{\mu}, \Sigma_S(\cdot)\right)$, again with a 
spatial kernel. Here, we set $\boldsymbol{\mu}=\mathbf{0}$, although the modeling can 
be carried out with any selection of $\boldsymbol{\mu}$. Standard practice would 
be to use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$.
Yet we have found that prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$.  Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\boldsymbol{\mu}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{aligned}
\end{equation}
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{aligned}
\]
In the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  In the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.
We considered setting the prior mean of $W$ to $X$ instead of $\mathbf{0}$
\citep[which would indicate stationarity apriori;][]{schmidt2003bayesian}, but found
it had a neglible effect on the model fits.

@@

While our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~\ref{eq:dgphm} into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\boldsymbol{\mu}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in \ref{sec:apdx_int_lik}. We then use elliptical slice
sampling \citep[ESS;][]{murray2010elliptical} to generate posterior samples of $W$.  
Proposals are generated from prior samples from Eq.~(\ref{eq:prior}), and acceptances 
involve the ratio of likelihoods from Eq.~\ref{eq:likelihood}.

We also embrace a fully-Bayesian treatment of unknown kernel hyperparameters 
$\{\tau^2, \theta_S, \theta_W\}$.  We prescale $X$ to $[0, 1]$
and $\bar{Y}$ to zero mean and unit variance.  On this scale, we adopt the default 
prior specifications provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
\textcolor{red}{(@Annie - from our email 5/29, can you comment on how we handle $\tau^2$/
cite Bobby's book/etc? Thanks!)}
%\textcolor{red}{Annie, please ensure the most up-to-date 
%prior specs are shown above and here. Here is what I found from fit\$settings when using 
%deepgp v1.1.3: $\theta_S \sim G(\alpha=1.2, \beta=1)$ and $\tau_S \propto 1/\tau_S^2$.}
We integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework with ESS
sampling of $W$ following the scheme of \citep{sauer2023active}.

@@

These results come from Bayes' Theorem, with details of this derivation provided 
in \ref{sec:apdx_SgivenY}. In our application, we find success setting 
$\boldsymbol{\mu}=\boldsymbol{0}$. With all posterior draws $S^{(t)}$, we compute
the average of these to obtain our posterior mean for a particular cosmology.
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.

>>>>

For each cosmology, Mira-Titan contains a batch of 18 simulated spectra: an inexpensive 
spectrum estimated from linear perturbation theory ($y_p$), 16 spectra estimated from 
lower resolutions simulations ($y_{\ell_r}, r \in \{1,\dots,16\}$), and one spectrum from 
a high resolution simulation ($y_h$). 
These spectra are represented as a function of wavenumber $k$.  Following 
\cite{moran2023mira}, we model on the emulation space 
$\mathcal{P}(k)=\log_{10}\left(\frac{k^{1.5}P(k)}{2\pi^2}\right)$, where $P(k)$ represents the
original spectra. Output is available for $n=351$ values of $k$, spanning
$0.001 \leq k \leq 5$, but each data type has unique wavenumber ranges
where it is deemed approximately unbiased. For $k<0.04$, only $y_p$ provides a reliable 
estimate of the true spectra. $y_{\ell_r}$ are valid for $0.04 \leq k \leq 0.25$, and 
$y_h$ is valid for $0.04 \leq k \leq 5$.  Figure \ref{fig:plot_data} shows an example 
of the output for a particular cosmology. When we focus on $k$ values where $y_{\ell_r}$ 
is valid (center panel), we see both low-resolution and high-resolution runs oscillate around 
a mean term, which is relatively smooth compared to these realizations. When this
mean term is subtracted, we will obtain a plot like the right panel in 
Figure \ref{fig:plot_data}.

@@

Here we describe our first contribution: the use of a Bayesian hierarchical model 
to estimate the underlying spectra and quantify uncertainty for a particular cosmology.
Let input $X$ represent the vector with elements $x_i = \log_{10}(k_i)$ for $i=1,\dots,n$.
In the previous section, we used lowercase $\bar{y}$ to denote 
the observed weighted average; now, we use $\bar{Y}$ to indicate the corresponding 
random variable within the statistical model.  Together, $X$ and $\bar{Y}$ comprise the 
training data for a particular cosmology. $\bar{Y}$ is typically pre-processed
based on domain expertise, such as that defined in Section \ref{sec:data}.
We will reserve details of domain-specific pre-processing for the CAMB and Mira-Titan suites
for Sections \ref{subsec:camb} and \ref{subsec:mira_fit}, respectively.

@@

Let $S$ represent the true (but unknown) underlying power matter spectrum for
a particular cosmology.  We consider $\bar{Y}$ a noisy realization of $S$.
Specifically, we assume $\bar{Y}$ is a random realization of a Gaussian 
process centered at $S$ with some covariance, i.e., 
$\bar{Y}\mid S \sim \mathrm{GP}(S, \Sigma_\varepsilon)$.  While previous works
have considered diagonal $\Sigma_\varepsilon$ \citep{moran2024dpc}, we find it
essential to account for the smoothly varying spatial dependence of the
low resolution runs (as visualized in Figure \ref{fig:plot_data}).  For
this, we embrace variations of the Mat\'ern kernel \citep{stein1999interpolation},
tailored to each scenario (more on this in later subsections).
We consider $\Sigma_\epsilon$ as known, given domain-matter expertise. (right)?
\textcolor{red}{(Do we actually tailor this?  I thought it was consistent across the board.)}

We next place a Gaussian process prior on $S$, i.e.,
$S\sim \mathrm{GP}\left(\boldsymbol{\mu}_S, \Sigma_S(\cdot)\right)$, again with a 
Mat\'ern kernel. Here, we set $\boldsymbol{\mu}_S=\mathbf{0}$, although our model
is applicable to any selection of $\boldsymbol{\mu}_S$. Standard practice would 
use locations $X$ as the inputs to kernel $\Sigma_S(\cdot)$, yet we have found that 
prior to lack sufficient flexibility for our motivating
application.  In the Mira-Titan setting, some nonstationarity is expected due to 
baryonic acoustic oscillations when $-2 \leq \log_{10}(k) \leq -1$. To account 
for this, we incorporate an additional latent Gaussian process, $W$, which accommodates
nonstationarity by warping $X$ via $W \sim \mathrm{GP}\left(\boldsymbol{\mu}_W, \Sigma_W(X)\right)$.
$W$ is also an $n$-vector whose entries $w_i$ correspond to warped versions of each $x_i$.
Again, we set $\boldsymbol{\mu}_W = \mathbf{0}$ without loss of generatlity (we have
found $\boldsymbol{\mu}_W = X$ works similarly).
Ultimately, this yields the following hierarchical model:
\begin{equation}\label{eq:dgphm}
\begin{aligned}
\bar{Y}\mid S,\Sigma_\varepsilon &\sim \mathrm{GP}(S,\Sigma_\varepsilon) \\
S\mid W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_S(W)\right) \\
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right).
\end{aligned}
\end{equation}
The functional composition of GPs in the latter two lines forms a ``deep Gaussian process'' 
\citep{damianou2013deep,dunlop2018deep}. Additional latent layers could be considered, 
although we find one is sufficient.

We use the Mat\`ern kernels,
\[
\begin{aligned}
&\Sigma_S(W)^{i,j} = \tau^2 \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_S}} + 
  \frac{5d^2}{3\theta_S}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_S}}\right)
\quad\textrm{for}\quad 
d=||w_i-w_j||^2, \;\;\textrm{and} \\
&\Sigma_W(X)^{i,j} = \left( 1 + \frac{\sqrt{5}d}{\sqrt{\theta_W}} + 
  \frac{5d^2}{3\theta_W}\right) \exp\left(-\frac{\sqrt{5}d}{\sqrt{\theta_W}}\right)
\quad\textrm{for}\quad 
d=||x_i-x_j||^2.
\end{aligned}
\]
In the $S$ layer, hyperparameters $\tau^2$ and $\theta_S$ govern the 
correlation scale and lengthscale respectively.  In the $W$ layer, we allow for a
unique lengthscale $\theta_W$ but enforce unit scale following \citet{sauer2023active}.

@@

Our inferential goal is to obtain the posterior distribution of 
$S\mid\bar{Y},\Sigma_\varepsilon$, but this requires inferring latent quantity $W$.  
We first integrate over $S$ to condense our hierarchical model of Eq.~(\ref{eq:dgphm}) into
\begin{align}
\label{eq:likelihood}
\bar{Y} \mid W, \Sigma_\varepsilon &\sim \textrm{GP}(\mathbf{0}, \Sigma_S(W) + \Sigma_\varepsilon) \\
\label{eq:prior}
W &\sim \mathrm{GP}\left(\mathbf{0}, \Sigma_W(X)\right)
\end{align}
Details are provided in \ref{sec:apdx_int_lik}. The incorporation of $\Sigma_\epsilon$ in the
Gaussian likelihood of this outer layer is an essential upgrade to previous DGP inferential schemes.
This covariance incorporates knowledge of the smoothly varying dynamics from which $\bar{Y}$ was
generated.

We embrace fully-Bayesian inference of latent $W$ and kernel hyperparameters $\{\tau^2, \theta_S, \theta_W\}$.
Specifically, we use elliptical slice sampling \citep[ESS;][]{murray2010elliptical} to 
generate posterior samples of $W$, using Eq.~(\ref{eq:prior}) for proposal generation and 
Eq.~(\ref{eq:likelihood}) for likelihood-based acceptance.  
Under a reference prior $\pi(\tau^2)\propto\frac{1}{\tau^2}$,
we integrate $\tau^2$ from the likelihood of Eq.~(\ref{eq:likelihood}).  The resulting marginal 
posterior is equivalent to that with the plug-in estimator 
$\hat{\tau}^2 = \dots$ \citep[][Chapter 5]{gramacy2020surrogates}.
\textcolor{red}{(With $\Sigma_\epsilon$ on there, we can't actually do this.. unless we divide
$\Sigma_\epsilon$ by $\tau^2$...)}.
For $\theta_S$ and $\theta_W$, after pre-scaling, we adopt the default prior specifications 
provided in the {\tt deepgp} {\sf R}-package \citep{deepgp}.
We then integrate Metropolis-Hastings sampling of these parameters in a Gibbs framework 
with the ESS sampling of $W$ \citep{sauer2023active}.

@@

These results come from Bayes' Theorem, with details of this derivation provided 
in \ref{sec:apdx_SgivenY}. With all posterior draws $S^{(t)}$, we compute
the average of these to obtain our posterior mean for a particular cosmology.
We denote our inference procedure for the hierarchical model of Eq.~\ref{eq:dgphm} as
``DGP.HM'' given its deep Gaussian process foundation with additional hierarchical modeling (HM)
structure to incorporate multiple functional outputs.
\textcolor{red}{(How do you get UQ though?)}
